---
title: "Exponential distribution"
author: "Bartek Skorulski"
date: "March 14th, 2014"
output: rmarkdown::tufte_handout
---

# Introduction

In this project you will investigate the exponential distribution. We ilustrate its 
properites via simpulations in R. We use the follwing packages:

```{r, message=FALSE, warning=FALSE, echo=FALSE}
require("knitr")
require("markdown")
require("ggplot2")
require("ggthemes")
require("data.table")
```
# Exponential distribtion

_Exponential distriubution_ is given by the formula:
\[f(x)=\lambda e^{-\lambda x},\]
for $\lambda>0$ where $x\geq 0$. Its
mean is $\lambda^{-1}$, and its variance is $\lambda^{-2}$. 

```{r, echo=FALSE, fig.margin = TRUE, fig.cap = "Exponental distribution with lambda equal to 0.2"}
x <- seq(0,20,0.1)
lambda <- 0.2
ggplot() + geom_path(aes(x,dexp(x,lambda)))+labs(x="",y="")+theme_tufte()
```


# Sample mean and variance

Now we see how the sample mean approximate the
theoretical one.

```{r, fig.height=1.5, fig.cap = "Sample means converging to the theoretical one."}
N <- 10000
set.seed(1110)
exps <- rexp(N,lambda)
means <- sapply(1:N, function(n) mean(exps[1:n]))
ggplot() + geom_line(aes(x=1:N, y=means)) + 
  geom_hline(yintercept=1/lambda, color="red") +
    labs(x="") + theme_tufte()
```
And now we see how the sample variance approximate the theoretical one.
```{r, fig.height=1.5, fig.cap = "Sample variances converging to the theoretical one."}
vars <- sapply(1:N, function(n) var(exps[1:n]))
ggplot()+geom_line(aes(2:N, vars[2:N]))+
  geom_hline(yintercept=1/lambda^2, color="red")+
    labs(x="", y="variances")+theme_tufte()
```

# Sum of independent random variable with exponential distribution

Exponential distribution is a special case of
_gamma distribution_
\[  f_{\alpha, \nu}(x) = \frac{1}{\Gamma(\nu)}\alpha^\nu x^{\nu-1}e^{-\alpha x}, \nu>0, x>0.\]
with $\alpha=\lambda$ and $\nu =1$. Let us also recall that:
\[  \Gamma(t) = \int_0^\infty x^{t-1}e^{-x}dx.\]

It is also worth noting that if $X_1$ and $X_2$ are two independent
random variable
that have gamma distribution respectively $f_{\alpha,\mu}$,
$f_{\alpha,\nu}$ with the same parameter $\alpha$, then
the sum $X_1+X_2$ has the distribution $f_{\alpha,\mu+\nu}$. Hence
sum $k$ of identicaly distributed random variables with expotential
distribribution is a gamma distribution $f_{\alpha, k}$. Moreover, if
$X$ is a Gamma distribution with a parametrs $\lambda, \nu$, then $X/k$ is
Gamma distribution with paramenter $k\lambda, \nu$.
Therefore, the density of $(X_1+\ldots+X_k)/k$ is $f_{k\lambda,k}$. So lets compare
Gamma distribution $f_{8, 40}$ that correspons to the distribution of the mean of 40 exponential samples with $\lambda=0.2$ and the normal distribution that should 
approximate this mean.

```{r, fig.margin=TRUE, fig.cap="Gamma and normal distribution"}
x <- seq(2,8,0.01)
k <- 40
e40.m <- 1/lambda
e40.sd <- 1/(lambda*sqrt(40))
ggplot()+geom_path(aes(x,dgamma(x, shape= k, rate = k*lambda), colour="blue"))+
    geom_path(aes(x=x, y = dnorm(x, e40.m, e40.sd), colour="red") )+
        labs(x="", y="")+
                 scale_colour_discrete(name  ="Distribution",
                                       labels=c("Normal", "Gamma"))+
  theme_tufte()

```

# Simulation

```{r, fig.fullwidth = TRUE, fig.height=2, message=FALSE, warning=FALSE, fig.cap="Histogram of 1000 sample means and corresponding normal and Gamma distribution."}
N <- 1000
means <- c()
set.seed(10)
for (n in 1:N) {
  exps <- rexp(k, lambda)
  means <- c(means,mean(exps))
}
e40.dt <- data.table(means=means)
x <- seq(e40.m-3.5*e40.sd,e40.m+3.5*e40.sd,e40.sd/10)
ggplot()+geom_histogram(data=e40.dt,aes(x=means, y=..density..),
                        fill="white", colour="black")+
    geom_path(aes(x=x, y = dnorm(x, e40.m, e40.sd),colour="red"))+theme_bw()+
        geom_path(aes(x,k*dgamma(k*x, shape= k, scale = 5), colour="blue"))+
            scale_colour_discrete(name  ="Distribution",
                                  labels=c("Simulation","Normal", "Gamma"))+
  theme_tufte()
```











